---
title: "7 More Data Vis"
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(IPEDS)
library(plotly)
admissions <- adm2020
```



## What this module adds
The last module used ggplot2 to make quick plots that helped us see the data. This module builds on that toward plots that ease our own interpretation or communicate clearly to someone else. We stay with the same ggplot2 package, but we add more of its functions to map several variables at once, layer geoms, add trend lines and reference lines, split a plot into small panels, control color and axes, and labeling everything so a reader can understand the plot without us standing next to them.

## ggplot2 extended
Every ggplot is built from layers and they can be added one at a time. In the last module we used the first few. Here is the fuller set, roughly in the order you may tend to add them. Importantly, every layer is not needed for every plot. A quick look might use only data, aesthetics, and a geom. A finished plot for a report can use more of them.


| Layer | What it controls | Example |
|:------|:-----------------|:--------|
| Data | The tibble being plotted | `ggplot(data, ...)` |
| Aesthetics | Which columns map to which visual properties | `aes(x = , y = , color = )` |
| Geoms | The shapes drawn | `geom_point()`, `geom_line()` |
| Stats | Stats a geom calculates before drawing | binning, smoothing |
| Scales | How values map to color, size, and axis positions | `scale_color` |
| Facets | Splitting the plot into panels by group | `facet_wrap(~ group)` |
| Coordinates | The plotting area and how it is viewed | `coord_cartesian()` |
| Theme | The overall look. Fonts, gridlines, legend | `theme_minimal()` |
| Labels | Titles, axis labels, and captions | `labs(title = )` |

: The layers of a ggplot. {#tbl-gg-layers}


## Follow along
The plots below are built from real data so you can run them yourself as you read. To follow along, load the tidyverse and the IPEDS package, then assign the admissions data to an object. If you have not installed these packages yet, install them once with `install.packages()`.

```r
# Load the packages (install first if needed)
library(tidyverse)
library(IPEDS)
library(plotly)

# Assign the admissions data to work with
admissions <- adm2020
```

## Building with `+` and `|>`
By now you have used two operators that both chain steps together, and it is worth being clear about when each one applies. Mixing them up is a common early error.

The pipe `|>` moves data from one dplyr verb to the next. It reads as "and then". Example: "Take the data, and then filter, and then summarize".

The plus `+` adds a layer to a ggplot. It reads as "and also". Example: "Start the plot, and also add points, and also add a trend line, and also add labels"

The two often appear together. You pipe data into `ggplot()` with `|>`, and from there you build the plot with `+`.

```{r}
#| label: pipe-plus-demo
#| warning: false
admissions |>                              # pipe the data in with |>
  filter(APPLCN > 30000) |>                  # dplyr steps use |>
  ggplot(aes(x = APPLCN, y = ADMSSN)) +     # switch to + once inside ggplot
  geom_point(alpha = 0.4)                   # add layers with +
```




## Mapping versus setting
The distinction between mapping and setting is important to understanding ggplot. To see it, we start with a plain plot and add color to it two different ways.

Here is a scatter with no color specified. Every point is ggplot's default black.

```{r}
#| label: plain-scatter
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN)) +
  geom_point(alpha = 0.2)
```




### Mapping color to a variable
To **map** an aesthetic means to connect it to a column so it varies with the data. Mapping goes *inside* `aes()`. Here we map color to whether an institution requires an admissions test, so the points take on different colors by group and a legend appears to explain them.

```{r}
#| label: map-color
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN, color = adm_tscores)) +
  geom_point(alpha = 0.4)
```

The color now carries information, showing a third variable on top of the two on the axes. ggplot chose the colors and built the legend for us because we mapped color to a column.





### Setting a constant color
To **set** an aesthetic means to fix it to one value that is the same for every point. Setting goes *outside* `aes()`, as an argument to the geom. Here every point is the same blue, and no legend appears because the color carries no information about the data.

```{r}
#| label: set-color
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN)) +
  geom_point(alpha = 0.4, color = "#799BA6")
```

In summary, if a visual property should change with the data, map it inside `aes()`. If it should be constant, set it outside `aes()` in the geom.

One related trap is worth seeing. If you put a constant *inside* `aes()`, it does not do what you might expect.

```{r}
#| label: aes-trap
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN)) +
  geom_point(aes(color = "blue"))
```

The points did not turn blue. Because `color = "blue"` sits inside `aes()`, ggplot treated the word "blue" as a category to map, gave it its own color, and added a legend labeled "blue". This is the classic sign that a constant ended up inside `aes()` by mistake.

### Global and local aesthetics
There is one more piece to where `aes()` goes. Aesthetics can live in two places, and the location decides which geoms they affect.

**Global**

An aesthetic placed inside `ggplot()` is **global**. Every geom in the plot inherits it. Here both the points and the smooth trend line are colored by group, because color is mapped globally.

```{r}
#| label: global-aes
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN, color = adm_tscores)) +
  geom_point(alpha = 0.4) +
  geom_smooth(se = FALSE)
```


**Local**

An aesthetic placed inside a single geom is **local**. Only that geom uses it. Here color is mapped only inside `geom_point()`, so the points are colored by group but the single smooth line is not, because it did not inherit the color mapping.

```{r}
#| label: local-aes
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN)) +
  geom_point(aes(color = adm_tscores), alpha = 0.4) +
  geom_smooth(se = FALSE)
```

The difference is visible in the trend lines. Global color gives one line per group, while local color leaves a single line over all the data. Deciding whether an aesthetic is global or local is how you control which parts of a plot respond to it.


## Building up a richer plot
So far we have mapped a single variable to color and seen how mapping and setting differ. From here we add more to a plot a piece at a time. Each addition is another layer joined with `+`. We will keep working with the admissions data and grow one plot as we go, so you can see each layer's effect on its own.


### Adding a shape dimension
Color is one aesthetic, and a plot can map more than one at a time. Shape is another. It draws each group's points with a different symbol, such as circles, triangles, and squares. Mapping shape works just like mapping color, inside `aes()`.

Here we map both color and shape to the same variable, the admissions test policy. Each group now differs in two ways at once, which can make groups easier to tell apart, and it helps a reader who cannot rely on color alone.

```{r}
#| label: map-shape
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.5)
```

The legend now shows both a color and a symbol for each group, since both aesthetics carry the same information.

Shape can also carry a *different* variable than color, letting one plot show more of the data at once. This is powerful, but it comes with a caution. The more a plot tries to show, the harder it can be to read. Being able to add another aesthetic does not always mean we should.

```{r}
#| label: shape-second-var
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = hs_gpa)) +
  geom_point(alpha = 0.5)
```

Here color shows the test policy and shape shows the high-school-GPA policy, so the plot carries four variables at once: two on the axes and two in the aesthetics. That is a lot to take in. A plot like this can be worth the density when a reader needs all of it, but often a cleaner plot that shows less communicates more.


### Zooming in on a range
Sometimes the interesting data sits in one part of the plot. Zooming into a a range can be helpful to examine and highlight them. There are two ways to zoom in ggplot2, and they behave differently in a way worth understanding.


#### Approach 1: coord_cartesian() for true zoom
`coord_cartesian()` sets the viewing window without changing the data. Points outside the window are simply off screen, but they are still there and still used in any calculation. This is a 'true zoom', like moving a magnifying glass over the plot.

```{r}
#| label: zoom-coord
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.6) +
  coord_cartesian(xlim = c(0, 25), ylim = c(0, 25))
```

Now the crowded region opens up. With fewer points overlapping, the colors and shapes do some explanatory work.This view sits at the very base of the full plot, the corner of the smallest institutions, where both applications and admissions are low. These are the many small schools that were packed into the bottom-left corner and invisible before we zoomed in.


#### Approach 2: scale limits, which drop data
Setting limits through the scales, with `xlim()` or `scale_x_continuous(limits = ...)`, looks similar but does something different. It removes any data outside the limits before the plot is drawn. The points beyond the range are not hidden, they are dropped, and anything computed from the data is recomputed on only what remains.

```{r}
#| label: zoom-scale
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.6) +
  geom_smooth(aes(group = 1), se = FALSE, color = "#C7522A") +
  xlim(0, 25) +
  ylim(0, 25)
```

Here we added a trend line to make the difference visible. Because the large institutions were removed before the line was fit, this line describes only the smaller institutions in the range, not the full dataset. You may also see a warning that rows were removed, which is ggplot telling you that data was dropped.


#### Which to use
The two look alike but answer different questions. Use `coord_cartesian()` when you want to look closely at part of the data while keeping every point and every calculation intact. Use scale limits when you genuinely want to exclude the data.


### Labeling the plot
So far the plot uses the raw column names for its axes, like `APPLCN` and `ADMSSN`, which a reader outside the data may not understand. The `labs()` layer sets readable labels for axis titles, plot titles, and captions. A useful habit is to put the plot's main takeaway in the subtitle, so a reader gets the point that you are trying to make.

```{r}
#| label: add-labels
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.6) +
  coord_cartesian(xlim = c(0, 25), ylim = c(0, 25)) +
  labs(
    title = "Applications and admissions at the smallest institutions",
    subtitle = "The smallest schools typically do not require or recommend admissions tests",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy",
    shape = "Test policy"
  )
```

The axes now read plainly, the title says what the plot is about, and the subtitle states the takeaway. Setting both `color` and `shape` labels to the same text also cleans up the legend, giving it one clear heading instead of the column name.


### Cleaning up the look with a theme
The default gray panel background is fine for quick looks, but a white background often reads better in a report. A theme controls the overall look of the plot: the background, the gridlines, the fonts. ggplot2 comes with several ready-made themes, and `theme_minimal()` gives a clean white background with light gridlines.

```{r}
#| label: add-theme
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = adm_tscores)) +
  geom_point() +
  coord_cartesian(xlim = c(0, 25), ylim = c(0, 25)) +
  labs(
    title = "Applications and admissions at the smallest institutions",
    subtitle = "The smallest schools typically do not require or recommend admissions tests",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy",
    shape = "Test policy"
  ) +
  theme_minimal()
```

The gray panel is gone, replaced by a clean white background that puts the focus on the points. This is the same plot as before, just easier to read, and it is starting to look like something you could drop into a report.



### Adjusting size and text
Two things can still be improved. The points are small, which makes the shapes hard to tell apart, and the default text is a little small for a report. We can fix both.

Point size is set in the geom, outside `aes()`, because we want every point the same larger size. We add `size = 2.5` to `geom_point()`.

Text sizes are controlled in the `theme()` layer. While `theme_minimal()` sets the overall look, `theme()` lets us adjust individual pieces, such as the size of the title, the axis labels, and the axis numbers. We add a `theme()` after `theme_minimal()` so our changes build on top of it.

```{r}
#| label: size-and-text
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.7, size = 3.5) +
  coord_cartesian(xlim = c(0, 25), ylim = c(0, 25)) +
  labs(
    title = "Applications and admissions at the smallest institutions",
    subtitle = "The smallest schools typically do not require or recommend admissions tests",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy",
    shape = "Test policy"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

The larger points make the shapes easy to distinguish, and the text sizes are more legible for a report or presentation. Notice the syntax inside `theme()` where each element of the plot has a name, such as `plot.title` or `axis.text`, and we adjust it with `element_text()`, which controls text properties like size and boldness. This is how ggplot2 gives fine control over nearly every part of a plot's appearance.See ggplot2 documentation to see the dozens of customizations available: <https://ggplot2.tidyverse.org/reference/theme.html>.


### Controlling the colors
So far ggplot2 has chosen the colors for us. Often that is fine, but sometimes we want to control that. For example, to use a colorblind-safe palette, to match a brand, or to pick colors that read well together. Color is set through a scale, and the scale for a mapped color aesthetic is one of the `scale_color_*()` functions.


#### A ready-made palette
The simplest control is to swap in a prebuilt palette. The viridis palettes are a popular choice because they are designed to be readable by people with color vision differences and to print well in grayscale. For a categorical variable like our test policy, we use `scale_color_brewer()` with the Set1 pallete. You can see the palettes here <https://ggplot2.tidyverse.org/reference/scale_brewer.html#palettes>.

```{r}
#| label: scale-viridis
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.7, size = 3.5) +
  coord_cartesian(xlim = c(0, 25), ylim = c(0, 25)) +
  labs(
    title = "Applications and admissions at the smallest institutions",
    subtitle = "The smallest schools typically do not require or recommend admissions tests",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy",
    shape = "Test policy"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal()
```




#### Setting each color by hand
When you want exact control, `scale_color_manual()` lets you assign a specific color to each group. You give it a set of values, one per group. Here we choose colors ourselves, including our site's slate and rust, so the plot matches the look we want.

```{r}
#| label: scale-manual
#| warning: false
ggplot(admissions, aes(x = APPLCN, y = ADMSSN,
                       color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.7, size = 3.5) +
  coord_cartesian(xlim = c(0, 25), ylim = c(0, 25)) +
  labs(
    title = "Applications and admissions at the smallest institutions",
    subtitle = "The smallest schools typically do not require or recommend admissions tests",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy",
    shape = "Test policy"
  ) +
  scale_color_manual(
    values = c(
      "Considered_but_not_required"      = "#C7522A",
      "Neither_required_nor_recommended" = "#799BA6",
      "Recommended"                      = "#6A8D73",
      "Required"                         = "#4A4A4A"
    )
  ) +
  theme_minimal()
```




### Marking a value on the plot
Beyond styling, we can put information directly on a plot to guide the reader such as lines marking a threshold or an average, and text labels explaining it. We build on the same plot, adding a reference line and an annotation.


#### A reference line
`geom_hline()` draws a horizontal line at a value we give it, and `geom_vline()` draws a vertical one. They are useful for marking a cutoff or an average. Here we compute the average admissions among these small institutions and draw a line at it in our orange accent color. This lets a reader see at a glance which schools sit above and below the average. 

We first filtered to the small institutions and stored them in `small_schools`, then computed the mean of their admissions as `mean_admit` and passed it to `geom_hline()`. The line comes straight from the data, so it stays correct if the data changes.

This occurs because `coord_cartesian()` leaves the entire dataset intact and only zooms in on a bit. These calculations ensure we aren't accounting for all schools when only the smallest ones are of interest in this moment.

```{r}
#| label: annotate-hline
#| warning: false
small_schools <- admissions |>
  filter(APPLCN <= 25, ADMSSN <= 25)

mean_admit <- mean(small_schools$ADMSSN, na.rm = TRUE)

ggplot(small_schools, aes(x = APPLCN, y = ADMSSN,
                          color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.7, size = 3.5) +
  geom_hline(yintercept = mean_admit, color = "#C7522A", linewidth = 1) +
  labs(
    title = "Applications and admissions at the smallest institutions",
    subtitle = "The orange line marks the average number of admissions",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy",
    shape = "Test policy"
  ) +
  scale_color_manual(
    values = c(
      "Considered_but_not_required"      = "#C7522A",
      "Neither_required_nor_recommended" = "#799BA6",
      "Recommended"                      = "#6A8D73",
      "Required"                         = "#4A4A4A"
    )
  ) +
  theme_minimal()
```


#### Adding a text label
A line shows *where* a value sits, but a label can say *what* it is. `annotate()` places text at coordinates we choose. Here we label the line so its meaning is clear on the plot itself.

```{r}
#| label: annotate-label
#| warning: false
ggplot(small_schools, aes(x = APPLCN, y = ADMSSN,
                          color = adm_tscores, shape = adm_tscores)) +
  geom_point(alpha = 0.7, size = 3.5) +
  geom_hline(yintercept = mean_admit, color = "#C7522A", linewidth = 1) +
  annotate("text",
           x = 5,
           y = mean_admit + 1.5,
           label = paste("Average:", round(mean_admit, 1)),
           color = "#C7522A",
           fontface = "bold") +
  labs(
    title = "Applications and admissions at the smallest institutions",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy",
    shape = "Test policy"
  ) +
  scale_color_manual(
    values = c(
      "Considered_but_not_required"      = "#C7522A",
      "Neither_required_nor_recommended" = "#799BA6",
      "Recommended"                      = "#6A8D73",
      "Required"                         = "#4A4A4A"
    )
  ) +
  theme_minimal()
```




### A bit of interactivity with plotly
Every plot so far has been a static fixed image. Sometimes it helps to let a reader explore a plot directly, hovering over points to read their exact values or zooming into a region. The plotly package can turn a ggplot into an interactive version with a single function, `ggplotly()`.

We build the plot as usual, save it to an object, then pass that object to `ggplotly()`. Here we use a clean version of our colored scatter.

```{r}
p <- ggplot(small_schools, aes(x = APPLCN, y = ADMSSN, color = adm_tscores)) +
  geom_point(alpha = 0.7, size = 3) +
  labs(
    title = "Applications and admissions at the smallest institutions",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy"
  ) +
  scale_color_manual(
    values = c(
      "Considered_but_not_required"      = "#C7522A",
      "Neither_required_nor_recommended" = "#799BA6",
      "Recommended"                      = "#6A8D73",
      "Required"                         = "#4A4A4A"
    )
  ) +
  theme_minimal()

ggplotly(p)
```


#### Exposing university names with plotly
We can also specify exactly what information we want plotly to display on hover.
Here we add the names of the universities to our small_schools dataset with a join.
In the aesthetic map, we define `text` that is later called by plotly by `ggplotly(p, tooltip = "text")`, rendering the value as a tooltip on hover.

```{r}
# The admissions data has IDs, not names, so we join names on first
# (the same dir_info2020 lookup from earlier).
small_named <- small_schools |>
  left_join(select(dir_info2020, INSTITUTION_ID, INSTITUTION),
            by = "INSTITUTION_ID")

p <- ggplot(small_named, aes(x = APPLCN, y = ADMSSN, color = adm_tscores,
                             text = INSTITUTION)) +
  geom_point(alpha = 0.7, size = 3) +
  labs(
    title = "Applications and admissions at the smallest institutions",
    x = "Applications received",
    y = "Students admitted",
    color = "Test policy"
  ) +
  scale_color_manual(
    values = c(
      "Considered_but_not_required"      = "#C7522A",
      "Neither_required_nor_recommended" = "#799BA6",
      "Recommended"                      = "#6A8D73",
      "Required"                         = "#4A4A4A"
    )
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```